---
title: "ps-5-markdown-plot-ryaradfar"
author: "Rya"
date: "2023-02-15"
output: html_document

```{r}
library(readr)
```
1 Load and check data (5pt)
You first task is to do a very simple data check:
  
1. (1pt) For solving the problems, and answering the questions, create a new rmarkdown docu-
ment with an appropriate title. See https://faculty.washington.edu/otoomet/info201-book/
r-markdown.html#r-markdown-rstudio-creating.

2. (2pt) Load data. How many rows/columns do we have?

```{r}
gapminder <- read_delim("gapminder.csv")
```
```{r}
gapminder %>% 
nrow()
```
```{r}
gapminder %>% 
ncol()
```

3. (2pt) Print a small sample of data. Does it look OK?

```{r}
gapminder %>% 
sample_n(10)

# Data looks good and has many missing values
```



2 Descriptive statistics (15pt)

1. (3pt) How many countries are there in the dataset? Analyze all three: iso3, iso2 and name.

```{r}
gapminder %>% 
summarize(n = n_distinct(iso3))

gapminder %>% 
summarize(n = n_distinct(iso2))

gapminder %>% 
summarize(n = n_distinct(name))

```
2. If you did this correctly, you saw that there are more names than iso-2 codes, and there are
even more iso3 -codes. What is going on? Can you find it out?
(a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that
correspond to more than one name? What are these countries?

```{r}
gapminder %>% 
group_by(iso2) %>% 
summarize(name_count = n_distinct(name)) %>% 
arrange(desc(name_count))

gapminder %>% 
filter(is.na(iso2)) %>% 
distinct(name)
```

(b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have
more than one iso3-code? What are these countries?

```{r}
gapminder %>% 
group_by(name) %>% 
summarize(name_count = n_distinct(iso3)) %>% 
arrange(desc(name_count)) 

gapminder %>% 
filter(is.na(name)) %>% 
distinct(iso3)
```

3. (2pt) What is the minimum and maximum year in these data?
```{r}
gapminder %>% 
  select(time) %>%
  filter(!is.na(time)) %>% 
  summarize(min(time), (max(time)))
```

Next, let’s analyze CO2 emissions.
1. (2pt) How many missing co2 emissions are there for each year? Analyze both missing CO2
and co2_PC. Which years have most missing data?

```{r}
gapminder %>%
  group_by(time) %>%
  summarize(missing_co2 = sum(is.na(co2)), missing_co2_PC = sum(is.na(co2_PC))) %>% 
  arrange(desc(missing_co2)) %>% 
  head(10)
```

2. (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few
more countries of your choice. Explain what do you see.

```{r}
gapminder %>% 
filter(iso3 %in% c("USA", "CHN", "IND", "NPL", "SEN", "KHM")) %>% 
ggplot(aes(time, co2, color = name)) +
geom_line() +
labs(x = "year", y = "co2 emissions", color = "country")

# China has the highest rate of co2 emissons per year, while Senegal has the lowest
# rate of co2 emission per year
```

3. (5pt) Now let’s analyze the CO2 emissions per capita (co2_PC ). Make a similar plot of the
same countries. What does this figure suggest?

```{r}
gapminder %>% 
filter(iso3 %in% c("USA", "CHN", "IND", "NPL", "SEN", "KHM")) %>% 
ggplot(aes(time, co2_PC, color = name)) +
geom_line() +
labs(x = "year", y = "co2 emissions pc", color = "country")

# This figure suggests that the US has the highest rate of co2 emissions per capita
```

4. (6pt) Compute average CO2 emissions per capita across the continents (assume region is the
same as continent). Comment what do you see.
Note: just compute averages over countries and ignore the fact that countries are of different
size.
Hint: Americas 2016 should be 4.80.

```{r}
average_co2_PC <- gapminder %>% 
filter(time == 1960 | time == 2016) %>%
filter(!is.na(co2_PC)) %>% 
filter(!is.na(region))%>% 
group_by(region, time) %>% 
summarize(avg_emissions_PC = mean((co2_PC)))

average_co2_PC

# The US has the highest average rate of cos emissions per capita in comparison
# to other continents
```

5. (7pt) Make a barplot where you show the previous results–average CO2 emissions per capita
across continents in 1960 and 2016.

```{r}

co2_1960_2016 <- filter(average_co2_PC, time %in% c(1960, 2016))
ggplot(co2_1960_2016, aes(x = "region", y = "co2_PC", fill = as.factor(time))) +
  geom_col(position = "dodge") +
  labs(title = "Average CO2 Emissions per Capita Across Continents",
       x = "region", y = "CO2 Emissions per Capita",
       fill = "time") +
  scale_x_discrete(labels = c("Africa", "Asia", "Europe", "Americas", "Oceania"))
```

6. Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per
capita) in 2019 for each continent? (Assume region is continent).

```{r}
co2_emmitters_large <- gapminder %>% 
                  filter(time == 2016) %>% 
                  filter(!is.na(region)) %>% 
                  filter(!is.na(co2_PC)) %>% 
                  group_by(region) %>%
                  select(region, time, co2_PC) %>%
                  arrange(desc(co2_PC)) %>% 
                  head(3)
co2_emmitters_large


co2_emmitters_small <- gapminder %>% 
                  filter(time == 2016) %>% 
                  filter(!is.na(region)) %>% 
                  filter(!is.na(co2_PC)) %>% 
                  group_by(region) %>%
                  select(region, time, co2_PC) %>%
                  arrange(desc(co2_PC)) %>% 
                  tail(3)
co2_emmitters_small
```

4 GDP per capita (50pt)

Let’s look at GDP per capita (GDP_PC ).
1. (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for
1960. Make the point size dependent on the country size, and color those according to the
continent. Feel free to adjust the plot in other ways to make it better.
Comment what do you see there.

```{r}
gapminder %>% 
  filter(time == 1960) %>% 
  ggplot(aes(GDP_PC, lifeExpectancy, color = region, size = totalPopulation)) +
  geom_point(alpha=0.5) +
  labs(x = "lifeExpectancy",
        y = "GDP_PC") +
   theme(text = element_text(size=13))
   
# Europe has high rates of life expectancy
# Asia and the Americas have high country sizes
```

2. (4pt) Make a similar plot, but this time use 2019 data only.

```{r}
gapminder %>% 
  filter(time == 2019) %>% 
  ggplot(aes(GDP_PC, lifeExpectancy, color = region, size = totalPopulation)) +
  geom_point(alpha=0.5) +
  labs(x = "GDP_PC",
        y = "lifeExpectancy") +
   theme(text = element_text(size=13))
```

3. (6pt) Compare these two plots and comment what do you see. How has world developed
through the last 60 years?

```{r}
# The world has developed to have a higher life expectancy rate 
# GDP per capita has also increased throughout the last 60 years
```

4. (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results
fit with what do you see on the figures?
Note: here as average I mean just average over countries, ignore the fact that countries are of
different size.

```{r}
gapminder %>% 
filter(time == 1960 | time == 2019) %>% 
filter(!is.na(lifeExpectancy)) %>% 
group_by(region, time) %>% 
summarize(mean = mean(lifeExpectancy))

# Results match what is shown in the figures
```

5. (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results
in the order of growth. Explain what do you see.

```{r}
gapminder %>% 
  filter(time >= 1961) %>% 
  group_by(region) %>% 
  arrange(name, time) %>% 
  mutate(avg_le = (lifeExpectancy - lag(lifeExpectancy)) / lag(lifeExpectancy)) %>% 
  summarize(avg_growth = mean(avg_le, na.rm = TRUE)) %>% 
  arrange(avg_growth)
  
# Oceania has the highest average life expectancy growth rate
```

6. (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both
histograms on the same graph, see how well you can do it!

```{r}
gapminder %>% 
  filter(time == 1960 | time == 2016) %>% 
  ggplot(aes(GDP_PC, fill = factor(time))) + 
  geom_histogram(position = "dodge") 
```

7. (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When
counting from top.)
Hint: check out the function rank()!
Hint2: 17 for 1960.

```{r}
gapminder %>% 
  filter(time == 1960 | time == 2016) %>% 
  filter(!is.na(lifeExpectancy)) %>% 
  filter(!is.na(region)) %>% 
  mutate(rank = rank(-lifeExpectancy)) %>% 
  filter(name == "United States of America")
```

8. (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a
bit. But we also have more countries in 2019–what about the relative rank divided by the
corresponding number of countries that have LE data in the corresponding year?
Hint: 0.0904 for 1960.

```{r}
gapminder %>% 
  filter(time == 1960) %>% 
  filter(!is.na(lifeExpectancy)) %>% 
  filter(!is.na(region)) %>% 
  mutate(le_rank = rank(desc(lifeExpectancy)), nc = n(), re_rank = nc/le_rank) %>% 
  filter(name == "United States of America")
```

Finally tell us how many hours did you spend on this PS.
```{r}
# About 10 hours
```